---
title: 'On zero-inflated permutation testing and some related problems: web material'
author: "Livio Finos and Fortunato Pesarin"
date: "28 July 2018"
output: 
  html_document: 
    toc: yes
---

This is the web material of the work **On zero-inflated permutation testing and some
related problems** by Livio Finos and Fortunato Pesarin.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(flip)
library(plyr)
library(knitr)
source("functions.R")
```

# Analysis of Medical Care data
The codes and the data are available [here](https://github.com/livioivil/flip_extra/tree/master/Zero-Inflated_Finos_Pesarin_2018)

## Load data
The data comes from:
<https://www.jstatsoft.org/index.php/jss/article/downloadSuppFile/v027i08/DebTrivedi.rda.zip>
A `Rdata` file is saved in this folder for your convenience.

Now, load the data and define the dataset to be used, for our convenience.
```{r load}
load("DebTrivedi.rda")

dati=DebTrivedi[,c("ofp","hosp","health","numchron","gender","school","privins")]
summary(dati)
```

## Fit parametric Hurdle model

```{r hurdle}
formula <- ofp ~ hosp + health + numchron + gender + school + privins
modelHurdle <- pscl::hurdle(formula = formula,
                            dist    = "negbin",
                            data    = DebTrivedi)
(out=summary(modelHurdle))

```

## Permutation based analysis


Before we proceed, let's compute the Maximum cardinality for each test:

```{r cardinality,warning=FALSE}
sapply(2:ncol(dati),compute_orbit_cardinality)
```

The number is very high, certainly enough for a permutation test.

```{r permutation}
# Depending on the number of permutations 'perms', the computation time can be very high, set EVAL=TRUE (here below) to run the analysis
EVAL=FALSE
perms=1000

dati$health=relevel(dati$health,"average")
dataX <- flip:::.getXY(ofp ~ hosp + health + numchron + gender + school + privins,
                         data=dati,statTest="permutation",rotationTest = FALSE)
nomefile=paste0("results_permutation_MC",perms,".Rdata")
if(EVAL){
  
  # the two response variables
  id_Ynot0=which(dataX$Y!=0)
  # Yno0=dataX$Y[dataX$Y!=0,]
  Y01=(dataX$Y!=0)*1
  # table(Y01)
  # Yno0_orig=Yno0
  Y01_orig=Y01
  
  lista_tested_var=list(2,3,4,5,6,7,8)
  set.seed(1)
  res_permT=llply(lista_tested_var,compute_test_stats_pvalues)
  save(res_permT,file=nomefile)
} else
load(file=nomefile)

## some make up of the results:
pvals=sapply(res_permT,t2p,obs.only = 1,tail=0)
pvals=c(pvals[2,],pvals[1,])

stats=unlist(sapply(res_permT,function(X)X[1,]))
stats=c(stats[2,],stats[1,])

out=new("flip.object")
out@res= data.frame(Test="z",Stat=stats,tail="><", p.value=pvals)
colnames(out@res)[4]="p-value"
rownames(out@res)=paste(rep(c("X|X>0 (Quantitative)","X>0 (Zeros)"),each=7),colnames(dataX$X)[-1],sep=":")
summary(out)
```




## Plots
An example for `gender(male)`. Change the number 5 in `res_permT[[5]]` - in the code below - to see other distributions of the test statistic

```{r plot}
out=new("flip.object",permT=res_permT[[5]][,2:1])
colnames(out@permT)=c("X|X>0 (Quantitative)","X>0 (Zeros)")
out@res=data.frame(T=matrix(res_permT[[5]][1,]))

par(mfrow=c(2,2))
plot(out)
hist(out[2],main=colnames(out@permT)[2])
hist(out[1],main=colnames(out@permT)[1])
```